---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#|",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tinkr

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/ropensci/tinkr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci/tinkr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/ropensci/tinkr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/ropensci/tinkr?branch=main)
<!-- badges: end -->

```{r child='man/rmd-fragments/goal.Rmd'} 
```  

## Use Cases

```{r child='man/rmd-fragments/use-cases.Rmd'} 
```

## Workflow

```{r child='man/rmd-fragments/workflow.Rmd'} 
```


## Installation

Wanna try the package and tell us what doesn't work yet? 

```r 
install.packages("tinkr", repos = "https://ropensci.r-universe.dev")
```

## Examples

### Markdown

```{r child='man/rmd-fragments/example-markdown-headers.Rmd'} 
```

### R Markdown

```{r child='man/rmd-fragments/example-rmarkdown-chunks.Rmd'} 
```

### Inserting new Markdown elements

```{r child='man/rmd-fragments/example-rmarkdown-adding-elements.Rmd'}
```

## Background / pre-requisites

```{r child='man/rmd-fragments/prereqs.Rmd'}
```


## Loss of Markdown style

### General principles and solution

```{r child='man/rmd-fragments/format-list.Rmd'}
```

  
### The special case of tables
  
```{r child='man/rmd-fragments/format-table.Rmd'}
```


### LaTeX equations

```{r child='man/rmd-fragments/format-latex.Rmd'}
```


## Meta

Please note that the 'tinkr' project is released with a [Contributor Code of Conduct](https://ropensci.org/code-of-conduct). By contributing to this project, you agree to abide by its terms.
