---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#|",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tinkr

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip) 
  [![R build status](https://github.com/ropenscilabs/tinkr/workflows/R-CMD-check/badge.svg)](https://github.com/ropenscilabs/tinkr/actions)
 [![Coverage status](https://codecov.io/gh/ropenscilabs/tinkr/branch/master/graph/badge.svg)](https://codecov.io/github/ropenscilabs/tinkr?branch=master)
  <!-- badges: end -->



The goal of tinkr is to convert (R)Markdown files to XML and back to allow their editing with `xml2` (XPath!) instead of numerous complicated regular expressions. Would you like to kknow more? [This is great intro if you are new to XPath](https://www.w3schools.com/xml/xpath_intro.asp) and [this is a good resource on XSLT for XML transformations](https://www.w3schools.com/xml/xsl_intro.asp).

## Use-Cases

Possible applications are R scripts using this and XPath in `xml2` to:

* change levels of headers, cf [this script](inst/scripts/roweb2_headers.R) and [this pull request to roweb2](https://github.com/ropensci/roweb2/pull/279)
* change chunk labels and options
* extract all runnable code (including inline code)
* insert arbitrary markdown elements 
* modify link URLs 
* your idea, feel free to suggest use cases!

## Workflow

Only the body of the (R) Markdown file is cast to XML, using the Commonmark specification via the [`commonmark` package](https://github.com/jeroen/commonmark). YAML metadata could be edited using the [`yaml` package](https://github.com/viking/r-yaml), which is not the goal of this package.

We have created an [R6 class](https://r6.r-lib.org/) object called **yarn** to store the representation of both the YAML and the XML data, both of which are accessible through the `$body` and `$yaml` elements. In addition, the namespace prefix is set to "md" in the `$ns` element.

You can perform XPath queries using the `$body` and `$ns` elements:

```{r}
library("tinkr")
library("xml2")
path <- system.file("extdata", "example1.md", package = "tinkr")
ex1 <- tinkr::yarn$new(path)
# find all ropensci.org blog links
xml_find_all(
  x = ex1$body, 
  xpath = ".//md:link[contains(@destination,'ropensci.org/blog')]", 
  ns = ex1$ns
)
```

## Installation

Wanna try the package and tell me what doesn't work? 

```r 
remotes::install_github("ropenscilabs/tinkr")
```


## Examples

### Markdown

This is a basic example. We read "example1.md", change all headers 3 to headers 1, and save it back to md. Because the {xml2} objects are [passed by reference](https://blog.penjee.com/wp-content/uploads/2015/02/pass-by-reference-vs-pass-by-value-animation.gif), manipulating them does not require reassignment. 


```{r}
library("magrittr")
library("tinkr")
# From Markdown to XML
path <- system.file("extdata", "example1.md", package = "tinkr")
# Level 3 header example:
cat(tail(readLines(path, 40)), sep = "\n")
ex1  <- tinkr::yarn$new(path)
# transform level 3 headers into level 1 headers
ex1$body %>%
  xml2::xml_find_all(xpath = ".//md:heading[@level='3']", ex1$ns) %>% 
  xml2::xml_set_attr("level", 1)

# Back to Markdown
tmp <- tempfile(fileext = "md")
ex1$write(tmp)
# Level three headers are now Level one:
cat(tail(readLines(tmp, 40)), sep = "\n")
unlink(tmp)
```

### R Markdown

For R Markdown files, to ease editing of chunk label and options, `to_xml` munges the chunk info into different attributes. E.g. below you see that `code_blocks` can have a `language`, `name`, `echo` attributes.

```{r}
path <- system.file("extdata", "example2.Rmd", package = "tinkr")
rmd <- tinkr::yarn$new(path)
rmd$body
```

Note that all of the features in {tinkr} work for both Markdown and R Markdown.

### Inserting new markdown elements

Inserting new nodes into the AST is surprisingly difficult if there is a default
namespace, so we have provided a method in the **yarn** object that will take
plain markdown and translate it to XML nodes and insert them into the document
for you. For example, you can add a new code block:

```{r new-block}
path <- system.file("extdata", "example2.Rmd", package = "tinkr")
rmd <- tinkr::yarn$new(path)
xml2::xml_find_first(rmd$body, ".//md:code_block", rmd$ns)
new_code <- c(
  "```{r xml-block, message = TRUE}",
  "message(\"this is a new chunk from {tinkr}\")",
  "```")
new_table <- data.frame(
  package = c("xml2", "xslt", "commonmark", "tinkr"),
  cool = TRUE
)
# Add chunk into document after the first chunk
rmd$add_md(new_code, where = 1L)
# Add a table after the second chunk:
rmd$add_md(knitr::kable(new_table), where = 2L)
# show the first 21 lines of modified document
rmd$head(21)
```

## Background / pre-requisites

If you are not closely following one of the examples provided, what background knowledge do you need before using tinkr?

* That XPath, a language for querying XML & HTML, exists, and [some basics](https://www.w3schools.com/xml/xpath_intro.asp).
* Basics of how [xml2 works](https://blog.r-hub.io/2020/01/22/mutable-api/#exposing-the-c-api-in-xml2): how to find, replace, remove nodes etc.
* How to use R6 classes... although reading the examples should help you get the gist.
* [XSLT](https://ropensci.org/blog/2017/01/10/xslt-release/) but only if you are not happy with our [default stylesheet](#general-principles-and-solution).

## Loss of Markdown style

### General principles and solution

The (R)md to XML to (R)md loop on which `tinkr` is based is slightly lossy because of Markdown syntax redundancy, so the loop from (R)md to R(md) via `to_xml` and `to_md` will be a bit lossy. For instance 

 * lists can be created with either "+", "-" or "*". When using `tinkr`, the (R)md after editing will only use "-" for lists. 
 
 * Links built like `[word][smallref]` and bottom `[smallref]: URL` become `[word](URL)`.
 
 * Characters are escaped (e.g. "[" when not for a link).

 * [x] GitHub tickboxes are preserved (only for `yarn` objects)
 
 * Block quotes lines all get ">" whereas in the input only the first could have a ">" at the beginning of the first line.
 
 * For tables see the next subsection.
    
  Such losses make your (R)md different, and the git diff a bit harder to parse, but should _not_ change the documents your (R)md is rendered to. If it does, report a bug in the issue tracker!
  
  A solution to not loose your Markdown style, e.g. your preferring "*" over "-" for lists is to tweak [our XSL stylesheet](inst/extdata/xml2md_gfm.xsl) and provide its filepath as `stylesheet_path` argument to `to_md`.
  
### The special case of tables
  
* Tables are supposed to remain/become pretty after a full loop `to_xml` + `to_md`. If you notice something amiss, e.g. too much space compared to what you were expecting, please open an issue.

### LaTeX equations

While markdown parsers like pandoc know what LaTeX is, commonmark does not, and
that means LaTeX equations will end up with extra markup due to commonmark's
desire to escape characters.

If you have LaTeX equations that use either `$` or `$$` to delimit them, you
can protect them from formatting changes with the `$protect_math()` method (for
users of the `yarn` object) or the `protect_math()` funciton (for those using
the output of `to_xml()`). Below is a demonstration using the `yarn` object:

```{r protect-math}
path <- system.file("extdata", "math-example.md", package = "tinkr")
math <- tinkr::yarn$new(path)
math$tail() # malformed
math$protect_math()$tail() # success!
```

Note, however, that there are a few caveats for this:

1. The dollar notation for inline math must be adjacent to the text. E.G. 
   `$\alpha$` is valid, but `$ \alpha$` and `$\alpha $` are not valid.
2. We do not currently have support for bracket notation
3. If you use a postfix dollar sign in your prose (e.g. BASIC commands or a 
   Burroughs-Wheeler Transformation demonstration), you must be sure to either
   use punctuation after the trailing dollar sign OR format the text as code.
   (i.e. `` `INKEY$` `` is good, but `INKEY$` by itself is not good and will be 
   interepreted as LaTeX code, throwing an error:
    ```{r, basic, error = TRUE}
    path <- system.file("extdata", "basic-math.md", package = "tinkr")
    math <- tinkr::yarn$new(path)
    math$head(15) # malformed
    math$protect_math() #error
    ```

## Meta

Please note that the 'tinkr' project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.
