While Markdown parsers like pandoc know what LaTeX is,
commonmark does not, 
and that means LaTeX equations will end up with extra markup due to commonmark's desire to escape characters.

However, if you have LaTeX equations that use either `$` or `$$` to delimit them, 
you can protect them from formatting changes with the `$protect_math()` method (for users of the `yarn` object) 
or the `protect_math()` function (for those using the output of `to_xml()`). 
Below is a demonstration using the `yarn` object:

```{r protect-math}
path <- system.file("extdata", "math-example.md", package = "tinkr")
math <- tinkr::yarn$new(path)
math$tail() # malformed
math$protect_math()$tail() # success!
```

Note, however, that there are a few caveats for this:

1. The dollar notation for inline math must be adjacent to the text. E.G. 
   `$\alpha$` is valid, but `$ \alpha$` and `$\alpha $` are not valid.
2. We do not currently have support for bracket notation
3. Use of `$` as currency will still work. There are a few caveats:
   1. The search for math is greedy, so currency may be extracted along with
      the math unless a markdown element separates them:
      ```{r}
      # no separator, math is detected at the beginning of the line.
      ex <- textConnection("$5 for a $\\pi$")
      tinkr::yarn$new(ex)$protect_math()$get_protected()
      # With an emph separator
      ex <- textConnection("$5 for _any_ $\\pi$")
      tinkr::yarn$new(ex)$protect_math()$get_protected()
      ```
   2. Postfix dollar signs (e.g. using BASIC code demostrations) should be
      protected with backtics, or it will cause an error.

